On the discovered Cancer Driving Nucleotides (CDNs) –Distributions across genes, cancer types and patients

Version 1

Abstract

Abstract

A central goal of cancer genomics is to identify, in each patient, all the cancer driving mutations. Among them, point mutations are referred to as Cancer Driving Nucleotides (CDNs), which recur in cancers. The companion study shows that the probability of i recurrent hits in n patients would decrease exponentially with i; hence, any mutation with i ≥ 3 hits in the TCGA database is a high-probability CDN. This study characterizes the 50∼150 CDNs identifiable for each cancer type of TCGA (while anticipating 10 times more undiscovered ones) as follows: i) CDNs tend to code for amino acids of divergent chemical properties. ii) At the genic level, far more CDNs (>5-fold) fall on non-canonical than canonical cancer driving genes (CDGs). Most undiscovered CDNs are expected to be on unknown CDGs. iii) CDNs tend to be more widely shared among cancer types than canonical CDGs, mainly because of the higher resolution at the nucleotide than the whole-gene level. iv) Most important, among the 50∼100 coding region mutations carried by a cancer patient, 5∼8 CDNs are expected but only 0∼2 CDNs have been identified at present. This low level of identification has hampered functional test and gene targeted therapy. We show that, by expanding the sample size to 105, most CDNs can be identified. Full CDN identification will then facilitate the design of patient-specific targeting against multiple CDN-harboring genes.

Introduction

Tumorigenesis in each patient is driven by mutations in the patient’s genome. Hence, a central goal of cancer genomics is to identify all driving mutations in each patient. This task is particularly challenging because each driving mutation is present in only a small fraction of patients. As the number of driver mutations in each patient has been estimated to be >5 (Armitage and Doll 1954; Bozic et al. 2010; Hanahan and Weinberg 2011; Belikov 2017; Anandakrishnan et al. 2019), the total number of driver mutations summed over all patients must be quite high.

This study, together with the companion paper (supplement File S2), are based on one simple premise: In the massively repeated evolution of cancers, any advantageous cancer-driving mutation should recur frequently, say i times in n patients. The converse that non-recurrent mutations are not advantageous is part of the same premise. We focus on point mutations, referred to as Cancer Driving Nucleotides (CDNs), and formulate the maximum of i (denoted i*) in n patients if mutations are not advantageous. For example, in the TCGA database with n generally in the range of 500∼1000, i* = 3. Hence, any point mutation with i ≥ 3 is a CDN. At present, a CDN would have a prevalence of 0.3% among cancer patients. If the sample size approaches 106, a CDN only needs to be prevalent at 5×10−5, the theoretical limit (supplement File S2).

Although there are many other driver mutations (e.g., fusion genes, chromosomal aberrations, epigenetic changes, etc.), CDNs should be sufficiently numerous and quantifiable to lead to innovations in functional tests and treatment strategies. Given the current sample sizes of various databases (Cerami et al. 2012; Weinstein et al. 2013; Tate et al. 2019; de Bruijn et al. 2023), each cancer type has yielded 50∼150 CDNs while the CDNs to be discovered should be at least 10 times more numerous. The number of CDNs currently observed in each patient is 0∼2 for most cancer types. This low-level of discovery has limited functional studies and hampered treatment strategies.

While we are proposing the scale-up of sample size to discover most CDNs, we now characterize CDNs that have been discovered. The main issues are the distributions of CDNs among genes, across cancer types and, most important, among patients. In this context, cancer driver genes (CDGs) would be a generic term. We shall use “canonical CDGs” (or conventional CDGs) for the driver genes in the union set of three commonly used lists (Bailey et al. 2018; Sondka et al. 2018; Martínez-Jiménez et al. 2020). In parallel, CDN-harboring genes, referred to “CDN genes”, constitute a new and expanded class of CDGs.

The first issue is that CDNs are not evenly distributed among genes. The canonical cancer drivers such as TP53, KRAS and EGFR tend to have many CDNs. However, the majority of CDNs, especially those yet-to-be-identified ones, may be rather evenly distributed with each gene harboring only 1∼2 CDNs. Hence, the number of genes with tumorigenic potential may be far larger than realized so far. The second issue is the distribution of CDNs and CDGs among cancer types. It is generally understood that the canonical CDGs are not widely shared among cancer types. However, much (but not all) of the presumed cancer-type specificity may be due to low statistical resolution at the genic level.

The third issue concerns the distribution of CDNs among patients. Clearly, the CDN load of a patient is crucial in diagnosis and treatment. However, the conventional diagnosis at the gene level may have two potential problems. One is that many CDNs do not fall in canonical CDGs as signals from one or two CDNs get diluted. Second, a canonical CDG, when mutated, may be mutated at a non-CDN site. In those patients, the said CDG does not drive tumorigenesis. We shall clarify the relationships between CDN mutations and genes that may or may not harbor them.

The characterizations of discovered CDNs are informative and offer a road map for expanding the CDN list. A complete CDN list for each cancer type will be most useful in functional test, diagnosis and treatment. We should note that, given the sharing of CDNs across cancer types, CDN lists in a subset of cancer types may be of general use across other cancer types.

Results

In molecular evolution, a gene under positive selection is recognized by its elevated evolutionary rate (Fig. 1A and 1C). There have been numerous methods for determining the extent of rate elevation (Li et al. 1985; Nei and Gojobori 1986; Yang and Swanson 2002; Lawrence et al. 2013; Martincorena et al. 2017; Sherman et al. 2022) and cancer evolution studies have adopted many of them. However, no model has been developed to take advantage of the massively repeated evolution of cancers (Fig. 1B), which happens in tens of millions of people at any time.

Mutations in organismal evolution vs. cancer evolution. (A, B) A hypothetical example of DNA sequence evolution in organism vs. in cancer with the same number of mutations. (C) Mutation distribution in two species in the organismal evolution of A. (D and E) Mutation distribution in cancer evolution among 10 sequences may have D and E patterns. (F) Another pattern of mutation distribution in cancer evolution with a recurrent site but shows too few total mutations. Mutations of (F) are CDNs missed in the conventional screens.

In the whole-gene analysis, Fig. 1C-E are identical, each with A : S = 10 : 1 where A and S denote nonsynonymous and synonymous mutations, respectively. However, the presence of a 4-hit site in Fig. 1E is far less likely to be neutral than Fig. 1C and 1D. Although ratio in Fig. 1F, A : S = 4 : 1, is statistically indistinguishable from the neutral ratio of about 2.5 : 1, Fig. 1F in fact has much more power to reject the neutral ratio than Fig. 1C and 1D. After all, the probability that multiple hits are at the same site in a big genome is obviously very small.

The analyses of CDNs across the whole genome

For the entire coding regions in the cancer genome data, we define Ai (or Si) as the number of nonsynonymous (or synonymous) sites that harbor a mutation with i recurrences. Table 1 presents the distribution of Ai and Si across the 12 cancer types with n > 300 (Weinstein et al. 2013).

For neutral mutations, we define i* as the threshold above which the expected numbers of Ai would be <1, i.e. E[Ai≥i∗] < 1, The corollary is that all Ai≥i∗ sites are advantageous CDNs. (Since Si is ∼ Ai/2.3, the same i* would apply to Si as well: E[Si≥i∗] < 1) As i* is a function of the number of patients (n), it is shown mathematically in the companion study (supplement File S2) that i* = 3 for n < 1000. Interestingly, while the E[Ai≥3] is < 1, the expected E[Ai≥4] is ≪ 1, in the order of 0.001. Hence, i* = 4 may be considered unnecessarily stringent.

We should note that this study is constrained by n < 1000 in the TCGA databases. (Databases with larger n’s are also used where the actual n’s are often uncertain.) At i* = 3, we could detect only a fraction (<10%; see below) of CDNs. Many more tumorigenic mutations may be found in the i = 1 or 2 classes although not every one of them is a CDN. Since these two classes of mutations are far more numerous, they should account for the bulk of CDNs to be discovered. Indeed, Table 1 shows 76 Ai≥3 CDN mutations per cancer type but 681 A2 and 56,648 A1 mutations in the lower recurrence groups. If n reaches 105∼6, most of the undiscovered CDNs in the A1 and A2 classes should be identified (see supplement File S2).

In Table 2, we estimate the proportion of the A1 and A2 mutations that are possible CDNs. The relationships of A3/S3 > A2/S2, A2/S2 > A1/S1, and A1/S1 > A0/S0 are almost always observed in Table 1 with 32 (3 × 8 + 2 × 4) out of 36 such relationships. The use of A/S ratios may still under-estimate the selective advantages of A1∼3 mutations because S1∼3 may have slight advantages as well (see supplement File S2). Assuming S1 is truly neutral, we use S0 to S1 as the basis to calculate the excess of A1∼3 in Table 2 where 35 of the 36 Obs(Ai) > Exp(Ai) relationships can be observed. The implication is that hundreds and, likely low thousands, of A1’s and A2’ should be CDNs whereas we have only confidently identified ∼76 strong CDNs, on average, for a cancer type. (Note that A1 excesses are less reliable since a 1% error in the calculation would mean 566 CDNs.)

CDNs and the amino acids affected

We now ask whether the amino acid changes associated with CDNs bear the signatures of positive selection. Amino acids that have divergent physico-chemical properties have been shown to be under strong selection, both positive and negative (Chen, He, et al. 2019; Chen, Lan, et al. 2019). We note that, in almost all cases in cancer evolution, when a codon is altered, only one nucleotide of the triplet codon is changed. Among the 190 amino acid (AA, 20×19/2) pairs, only 75 of the pairs differ by one bp (Tang et al. 2004). For example, Pro (CCN) and Ala (GCN) may differ by only one bp but Pro and Gly (GGN) must differ by at least 2 bp. These 75 AA changes, referred to as the elementary AA changes (Grantham 1974; Li et al. 1985; Yang et al. 2003; Meyer et al. 2021), account for almost all AA substitutions in somatic evolution.

In a series of studies (Tang et al. 2004; Chen, He, et al. 2019; Chen, Lan, et al. 2019), we have defined the physico-chemical distances between AAs of the 75 elementary pairs as ΔUi, where i = 1 to 75. ΔUi reflects 47 measures of AA differences including hydrophobicity, size, charge etc. and ranges between 0 and 1. The most similar pair, Ser and Thr, has ΔUi = 0 and the most dissimilar pair is Asp and Try with ΔUi = 1. These studies show that ΔUi is a strong determinant of the evolutionary rates of DNA sequences and that large-step changes (i.e., large ΔUi’s) are more acutely “recognized” by natural selection. These large-step changes are either highly deleterious or highly advantageous. Most strikingly, advantageous mutations are enriched with AA pairs of ΔUi > 0.8 (Chen, He, et al. 2019).

To analyze the properties of CDNs, we choose 6 cancer types from Table 1 that have the largest sample sizes (n > 500) but leap over kidney since kidney cancers have unusually low CDN counts. In Fig. 2, we divide the CDNs into groups according to the number of recurrences, i. CDNs of similar i’s are merged into the same group in the descending order of i, until there are at least 10 CDNs in the group. The 6 cancer types show two clear trends: 1) the proportion of CDNs with ΔUi > 0.8 (red color segments) increases in groups with higher recurrences; 2) in contrast, the proportion of CDNs with ΔUi < 0.4 (green segments) decreases as recurrences increase. These two trends would mean that highly recurrent CDNs tend to involve larger AA distances (ΔUi > 0.8) and similar AAs tend not to manifest strong fitness increases. In general, CDNs alter amino acids in ways that expose the changes to strong selection.

ΔUi analysis across 6 cancer types. ΔUi, ranging between 0 and 1 (Tang et al. 2004; Chen, He, et al. 2019), is a measure of physico-chemical differences among the 20 amino acids (see the text). The most similar amino acids have ΔUi near 0 and the most dissimilar ones have ΔUi near 1. Each panel corresponds to one cancer type, with horizontal bar represents ΔUi distribution of each recurrence group. The numbers on the left of the panel are i values and on the right are the number of sites. Note that the proportion of dark red segments increases as i increases. This figure shows that mutations at high recurrence sites (larger i’s) code for amino acids that are chemically very different from the wild type.

CDNs in relation to the genes harboring them

We shall use the term “CDN genes” for genes having at least one CDN site. Since CDN genes contribute to tumorigenesis when harboring a CDN mutation, they should be considered cancer drivers as well. CDN genes have two desirable qualities for recognition as driver genes. First, CDNs are straightforward and unambiguous to define (e.g., i ≥ 3 for n < 1000). In the literature, there have been multiple definitions of cancer driver genes (Reimand and Bader 2013; Porta-Pardo and Godzik 2014; Mularoni et al. 2016; Arnedo-Pac et al. 2019), resulting in only modest overlaps among cancer gene lists (see Fig. S3). Second, the evolutionary fitness of CDN, and hence the tumorigenic potentials of CDN genes, can be computed (supplement File S1).

We now present the analyses of CDN genes, using the same 6 cancers of Fig. 2. Two types of CDN genes are shown in Table 3. Type I genes fulfill the conventional criterion of fast evolution with the whole-gene Ka/Ks (or dN/dS) significantly larger than 1 (Martincorena et al. 2017). Averaged across cancer types, Type I overlaps by 95.7% with the canonical CDG list, which is the union of three popular lists (Bailey et al. 2018; Sondka et al. 2018; Martínez-Jiménez et al. 2020). Type I genes are mostly well-known canonical CDGs (e.g., TP53, PIK3CA, and EGFR).

Type II (CDN genes) is the new class of cancer driver genes. These genes have CDNs but do not meet the conventional criteria of whole-gene analysis. Obviously, if a gene has only one or two CDNs plus some sporadic hits, the whole-gene Ka/Ks would not be significantly greater than 1. As shown in Table 3, over 80% of CDN genes have only 1∼2 CDN sites. The salient result is that Type II genes outnumber Type I genes by a ratio of 5 : 1 (229 : 45, column 8, Table 3). Furthermore, Type II genes overlap with the canonical CDG list by only 23%.

Type II genes represent a new class of cancer drivers that concentrate their tumorigenic strength on a small number CDN sites. They have been missed by the conventional whole-gene definition of cancer drivers. One such example is the FGFR3 gene in lung cancer. This gene of 809 codons has only 8 hits, among which one is a CDN (i = 3) in lung cancer. It is noticed solely for this CDN. In the supplemental text, we briefly annotate these new cancer driver genes for comparisons with the canonical driver genes (supplement File S1). Possible functional tests in the future can be found in Discussion.

We now briefly discuss the driver genes listed in previous studies as shown at the lower part of Table 3 (Bailey et al. 2018; Sondka et al. 2018; Martínez-Jiménez et al. 2020). From the total number of CDGs listed, it is clear that the overlaps are limited. As analyzed before (Wu et al. 2016), conventional gene lists overlap mainly by a core set of high Ka/Ks genes. This core set has not changed much as various criteria such the replication timing, expression profiles, epigenetic features are introduced. These criteria are the reasons for the many CDGs recognized by only a small subset of CDG lists. CDN genes, in contrast, can be objectively defined as CDN mutations (i recurrences in n samples) themselves are unambiguous.

Variation in CDN number and tumorigenic contribution among genes

By and large, the distribution of CDNs among genes is very uneven. Fig. 3A shows 10 genes with at least 6 CDNs whereas 87 genes have only one CDN. Two genes stand out for the number of CDNs they harbor, TP53 and PIK3CA, which also happen to be the only genes mutated in >15% of all cancer patients surveyed (Kandoth et al. 2013). Clearly, the prevalence of mutations in a gene is a function of the number of strong CDNs it harbors.

Distribution of CDNs among genes. (A) Out of 119 CDN-carrying genes (red bars), 87 have only one CDN. For the rest, TP53 possesses the most CDNs with three others having more than 10 CDNs. (B) CDN number in TP53 among patients. The dark bar represents the observed patient number with corresponding CDNs of the X-axis. The grey bar shows the expected patient distribution. Clearly, TP53 only needs to contribute one CDN to drive tumorigenesis. Hence, TP53 (and other canonical driver genes; see text), while prevalent, does not contribute disproportionately to the tumorigenesis of each patient.

Although a small number of genes have unusually high number of CDNs, these genes may not drive the tumorigenesis in proportion to their CDN numbers in individual patients. Fig. 3B shows the number of CDN mutations on TP53 that occur in any single patient. Usually, only one CDN change is observed in a patient whereas 2 or 3 CDN mutations are expected. It thus appears that CDNs on the same genes are redundant in their tumorigenic effects such that the second hit may not yield additional advantages. This pattern of disproportionally lower contribution by CDN-rich genes is true in other genes such as EGFR and KRAS. Consequently, the large number of genes with only 1 or 2 CDN sites are disproportionately important in driving the tumorigenesis of individual patients.

CDNs in relation to the cancer types - The pan-cancer properties

In the current literature, cancer driver genes (however they are defined) generally meet the statistical criteria for driver genes in only one or a few cancer types. However, genes may in fact contribute to tumorigenesis but are insufficiently prevalent to meet the statistical requirements for CDGs. Many genes are indeed marginally qualified as drivers in some tissues and barely miss the statistical cutoff in others. To see if genes that drive tumorigenesis in multiple tissues are more common than currently understood, we need to raise the sensitivity of cancer driver detection. Thus, CDNs may provide the resolution.

To test the pan-cancer driving capacity of CDNs, we define imax as the largest i values among the 12 cancer types for each CDN. The number of cancer types where the said mutation can be detected (i.e., i > 0) is designated NC12. Fig. 4 presents the relationship between the observed NC12 of each CDN against imax of that CDN. Clearly, many CDNs are observed in multiple cancer types (NC12 > 3), even though they do not qualify as a driver gene in all but a single cancer type. It happens frequently when a mutation has i > 3 in one cancer type but has i < 3 in others. One extreme example is C394 and G395 in IDH1. In CNS, both sites show i ≫ 3, while in 6 other cancer types (lung, breast, large intestine, prostate, urinary tract, liver), their hits are i < 3 but > 0. Conditional on a specific site informed by a cancer type, a mutation in another cancer type should be very unlikely if the mutation is not tumorigenic in multiple tissues. Hence, the pattern in Fig. 4 is interpreted to be drivers in multiple cancer types, but with varying statistical strength.

Sharing of CDNs across cancer types. The X-axis shows imax, which is the largest i a CDN reaches among the 12 cancer types. The Y axis shows the number of cancer types whre the mutation also occurs. Each dot is a CDN and the number of dots in the cloud is given. The blue and red dots denote, respectively, mutations classified as a CDN in one or multiple cancer types. Grey dots are non-CDNs. The table in the lower panel summarizes the number of sites and the number of genes harboring these sites.

Examining Fig. 4 more carefully, we could see that CDNs with a larger imax in one cancer type are more likely to be identified as CDNs in multiple cancer types (red dots, r = 0.97, p = 9.23×10−5, Pearson’s correlation test). Of 22 sites with imax > 20, 15 are identified as CDNs (i ≥ 3) in multiple cancer types, with a median NC12 of 9. On the opposite end, 2 CDNs with imax > 20 are observed in only one cancer type (EGFR: T2573 in lung and FGFR2: C755 in endometrium cancer). The bimodal pattern suggests that a few cancer driver mutations are tissue specific whereas most others appear to have pan-cancer driving potentials.

To conclude, when a driver is observed in more than one cancer type, it is often a cancer driver in many others, but insufficiently powerful to meet the statistical criteria for driver mutations. This pan-cancer property can be seen at the higher resolution of CDN, but is often missed at the whole-gene level. Cancers of the same tissue in different patients, often reported to have divergent mutation profiles (Nik-Zainal et al. 2012; Roberts and Gordenin 2014), should be a good test of this hypothesis.

CDNs in relation to individual patients and therapeutic strategies

In previous sections, the focus is on the population of cancer patients; for example, how many in the patient population have certain mutations. We now direct the attention to individual patients. It would be necessary to pinpoint the CDN mutations in each patient in order to delineate the specific evolutionary path and to devise the treatment strategy. We shall first address the cancer driving power of CDN vs. non-CDN mutations in the same gene.

1) Efficacy of targeted therapy against CDNs vs non-CDNs

In general, a patient would have many point mutations, only a few of which are strong CDNs. We may ask whether most mutations on the canonical genes, such as EGFR, are CDNs. Presumably, synonymous, and likely many nonsynonymous, mutations on canonical genes may not be CDNs. It would be logical to hypothesize that patients whose EGFR has a CDN mutation (Group1 patients) should benefit from the gene-targeted therapy more than patients with a non-CDN mutation on the same gene (Group2 patients). In the second group, EFGR may be a non-driver of tumorigenesis.

Published data (AACR Project GENIE Consortium 2017; Choudhury et al. 2023) are re-analyzed as shown in Fig. 5. The hypothesis that patients of Group2 would not benefit as much as those of Group1 is supported by the analysis. This pattern further strengthens the underlying assumption that non-CDN mutations, even on canonical genes, are not cancer drivers.

Survival analysis of non-small cell lung cancer (NSCLC) patients based on EGFR mutation status. Patient data were retrieved from the GENIE database (https://genie.cbioportal.org/) and stratified into three groups based on EGFR mutation profiles: Group1 comprises patients with EGFR CDN mutations; Group2 includes patients with nonsynonymous mutations in EGFR that are not CDNs; The EGFRWT group consists of patients with no EGFR mutations (see methods). Patients of Group1 and Group2 received EGFR-targeted therapies in accordance with the guidelines for managing EGFR mutant NSCLC (Passaro et al. 2022; Choudhury et al. 2023). Survival analysis using the Kaplan-Meier method revealed a significantly higher survival rate for Group1 patients compared to Group2 and the EGFRWT group (p < 0.001).

2) Number of CDNs in each patient

We postulate that a full set of CDNs should be able to inform about the cause of each cancer as well as the design of gene-targeted therapy. In Table 4, the known CDNs based on TCGA are tallied. Note that only a few CDNs fall on the canonical driver genes whereas most CDNs fall on the non-conventional ones.

In most cancer types, 10%∼30% of patients, shown in the n0 row of Table 4, have no known CDNs (and >50% among breast cancer patients). Hence, the current practice is to rely on missense mutations, regardless of CDNs or non-CDNs, on the canonical genes. The CDN column vs. the gene column in Table 4 address this issue. For example, the CDN column suggests that 33% of lung cancer patients (the n0 row) would not respond well to gene-targeted therapy whereas the gene column show only 5.3%. The difference is due to a higher, and likely inflated, detection rate of candidate drivers in the gene column. We suggest that patients who have a non-CDN mutation on a driver gene would not respond to the targeted therapy against that gene, as demonstrated in Fig. 5. In the above example, 27.7% (33%∼5.3%) of patients may be subjected to the targeted treatment but may not respond well.

3) Prevalence vs. potency of CDN-bearing genes in driving tumorigenesis

The last question is the relationship between mutation prevalence and tumorigenic strength (or potency) among CDN-bearing genes. For example, when a patient is diagnosed to have 5 CDNs in 5 genes, what may be their relative contributions to the tumorigenesis? Are they equally valid candidates for targeted therapy? It would seem logical that canonical CDGs with many CDNs should be the targets. However, because these genes would contribute at most one CDN to the tumorigenesis (Fig. 3B), targeting a high prevalence gene may not yield more benefits to the patients than targeting a low prevalence gene that has a CDN.

The implication is that prevalence and potency of CDNs may not be strongly correlated. Some genes may be prevalently mutated in the patient population but, in each affected patient, these genes may not be more potent than the less prevalent genes with a CDN mutation. Potency can be tested in vitro by gene editing or in vivo by targeting treatment. In this interpretation, targeting a CDN of low prevalence (say, i = 3) may be as effective in treatment as targeting a high prevalence CDN with i = 20. The model and Table 5 present this hypothesis based on cancer hallmarks.

The hallmarks of cancer were first proposed by (Hanahan and Weinberg 2000) with several updates (Hanahan and Weinberg 2011; Hanahan 2022). Each hallmark is a cancer phenotype shown in Table 5 that lists the number of genes involved in each particular hallmark (see Methods). While each hallmark may be associated with a number of genes, many genes are also involved in multiple hallmarks. As even the highly prevalent genes would usually have at most one mutation in each patient, we assume that each gene is associated with one hallmark in each patient.

Suppose that tumorigenesis requires a mutation in most (but perhaps not all) of the hallmarks, then the number of mutation combinations would be the product of all numbers in the corresponding column. For breast cancer, it would be 8 × 12 × 4 …. × 11 × 2 ∼ 1.7 × 1011. In other words, the possible mutation combinations that can drive breast cancer is over a billion. Hence, two breast cancers are unlikely to have the same set of CDGs or CDNs. In this view, the prevalence of a gene would be inversely proportional to the hallmark gene number. For example, genes of “invasion and metastasis” in breast cancer would have a prevalence of < 1/52. In contrast, the potency in tumorigenesis should depend on the hallmark phenotype itself, and independent of gene number for that hallmark. In this example, each gene of “invasion and metastasis” may be lowly prevalent, but could also be highly potent in each patient.

In short, the prevalence and potency of CDNs may be poorly correlated. The hypothesis can be functionally tested (by gene-editing in vitro or targeting treatment in vivo) in conjunction with the data on the attraction (i.e., co-occurrences) vs repulsion (lack of co-occurrences) of CDNs.

Discussion

The companion study presents the theory that computes the limit of recurrences (i/n, i times in n patients) of reachable by neutral mutations. Above the cutoff (e.g., 3/1000), a recurrent mutation is deemed an advantageous CDN. At present, the power of CDN analysis is hampered by the still small sample sizes, generally between 300 and 3000. We show that, when n reaches 105, a mutation only has to recur 12 times to be shown as a CDN, i.e., 25 times more sensitive than 3/1000. In short, nearly all CDNs should be discovered with n ≥ 105.

In this study, we apply the theory on existing data to characterize the discovered CDNs. Based on the TCGA data, this study concludes that each cancer patient carries only 1∼2 CDNs, whereas 6∼10 drivers are usually hypothesized to be present in each cancer genome (Hanahan and Weinberg 2011; Vogelstein et al. 2013; Campbell et al. 2020). This deficit signifies the current incomplete understanding of cancer driving potentials. Across patients of the same cancer type, about 50 to 150 CDNs have been discovered for each cancer type, representing perhaps only 10% of all possible CDNs. Given a complete set of CDNs, it should be possible to delineate the path of tumor evolution for each individual patient.

Direct functional test of CDNs would be to introduce putative cancer-driving mutations and observe the evolution of tumors. Such a task of introducing multiple mutations that are collectively needed to drive tumorigenesis has been done only recently, and only for the best-known cancer driving mutations (Ortmann et al. 2015; Takeda et al. 2015; Hodis et al. 2022). In most tumors, the correct combination of mutations needed is not known. Clearly, CDNs, with their strong tumorigenic strength, are suitable candidates.

Many CDNs in a patient may not fall on conventional CDGs, whereas these conventional CDGs may have passenger or weak mutations. Therefore, the efforts in gene-targeting therapy may well be shifted to the CDN-harboring genes. Given a complete set of CDNs, many more driver genes can be identified. Since many driver genes cannot be targeted for biological or technical reasons (Dang et al. 2017; Danesi et al. 2021; Waarts et al. 2022), a large set of CDGs will be desirable. The goal is that each cancer patient would have at least one targetable CDG driven by its CDN. In fact, it would be most beneficial if patients can have multiple targetable CDGs. In that case, the probability that resistant mutations eluding multiple targeting drugs should be diminished.

Another interesting insight revealed may be the distributions of CDNs across different cancer types. It suggests that CDNs previously identified to be cancer-specific may have pan-cancer effects. Such recurrences across cancer types are likely selection-driven but, in many cancer types, the strength is not sufficient to rise above the statistical threshold. While tumorigenesis in the same tissue type is a process of massively repeated evolution, cancer evolution in different tissues represents massive parallel evolution. The parallelism represents processes driven by the same selective forces (for cell proliferation) under different ecological conditions.

CDNs can also be used in cancer screening with the advantage of efficiency as the targeted mutations are fewer. Being efficient, the false negative rate should be lower too. Most interesting of all, the false positive rate should be far lower than the gene-based screen which often shows a false positive rate of >50% (supplement File S1).

Cancer evolution falls within the realm of ultra-microevolution (Wu et al. 2016). The repeated evolution addresses the single most severe criticism of evolutionary studies, namely all evolutionary events have a sample size of one. Individual advantageous mutations should repeat themselves as CDNs. Mutations that have been favored by positive selection in each individual patient are also the most fruitful targets in cancer therapy. The two studies thus unite evolutionary biology and cancer medicine.

eLife Assessment

This valuable study is a companion to a paper introducing a theoretical framework and methodology for identifying Cancer Driving Nucleotides (CDNs). While the evidence that recurrent SNVs or CDNs are common in true cancer driver genes is solid, the evidence that many more undiscovered cancer driver mutations will have CDNs, and that this approach could identify these undiscovered driver genes with about 100,000 samples, is limited.

Reviewer #1 (Public Review):

The study investigates Cancer Driving Nucleotides (CDNs) using the TCGA database, finding that these recurring point mutations could greatly enhance our understanding of cancer genomics and improve personalized treatment strategies. Despite identifying 50-150 CDNs per cancer type, the research reveals that a significant number remain undiscovered, limiting current therapeutic applications, and underscoring the need for further larger-scale research.

Strengths:

The study provides a detailed examination of cancer-driving mutations at the nucleotide level, offering a more precise understanding than traditional gene-level analyses. The authors found a significant number of CDNs remain undiscovered, with only 0-2 identified per patient out of an expected 5-8, indicating that many important mutations are still missing. The study indicated that identifying more CDNs could potentially significantly impact the development of personalized cancer therapies, improving patient outcomes.

Weaknesses:

The study is constrained by relatively small sample sizes for each cancer type, which reduces the statistical power and robustness of the findings. ICGC and other large-scale WGS datasets are publicly available but were not included in this study.

To be able to identify rare driver mutations, more samples are needed to improve the statistical power, which is well-known in cancer research.

The challenges in direct functional testing of CDNs due to the complexity of tumor evolution and unknown mutation combinations limit the practical applicability of the findings.

The QC of the TCGA data was not very strict, i.e, "patients with more than 3000 coding region point mutations were filtered out as potential hypermutator phenotypes", it would be better to remove patients beyond +/- 3*S.D from the mean number of mutations for each cancer type. Given some point mutations with >3 hits in the TCGA dataset, they were just false positive mutation callings, particularly in the large repeat regions in the human genome.

The codes for the statistical calculation (i.e., calculation of Ai_e, et al) are not publicly available, which makes the findings hard to be replicated.

Reviewer #2 (Public Review):

Summary:

The study proposes that many cancer driver mutations are not yet identified but could be identified if they harbor recurrent SNVs. The paper leverages the analysis from Paper #1 that used quantitative analysis to demonstrate that SNVs or CDNs seen 3 or more times are more likely to occur due to selection (ie a driver mutation) than they are to occur by chance or random mutation.

Strengths:

Empirically, mutation frequency is an excellent marker of a driver gene because canonical driver mutations typically have recurrent SNVs. Using the TCGA database, the paper illustrates that CDNs can identify canonical driver mutations (Figure 3) and that most CDNs are likely to disrupt protein function (Figure 2). In addition, CDNs can be shared between cancer types (Figure 4).

Weaknesses:

Driver alteration validation is difficult, with disagreements on what defines a driver mutation, and how many driver mutations are present in a cancer. The value proposed by the authors is that the identification of all driver genes can facilitate the design of patient-specific targeting therapies, but most targeted therapies are already directed towards known driver genes. There is an incomplete discussion of oncogenes (where activating mutations tend to target a single amino acid or repeat) and tumor suppressor genes (where inactivating mutations may be more spread across the gene). Other alterations (epigenetic, indels, translocations, CNVs) would be missed by this type of analysis.

The method could be more valuable when applied to the noncoding genome, where driver mutations in promoters or enhancers are relatively rare, or as yet to be discovered. Increasingly more cancers have had whole genome sequencing. Compared to WES, criteria for driver mutations in noncoding regions are less clear, and this method could potentially provide new noncoding driver CDNs. Observing the same mutation in more than one cancer specimen is empirically unusual, and the authors provide a solid quantitative analysis that indicates many recurrent mutations are likely to be cancer-driver mutations.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment

This valuable study is a companion to a paper introducing a theoretical framework and methodology for identifying Cancer Driving Nucleotides (CDNs). While the evidence that recurrent SNVs or CDNs are common in true cancer driver genes is solid, the evidence that many more undiscovered cancer driver mutations will have CDNs, and that this approach could identify these undiscovered driver genes with about 100,000 samples, is limited.

Same criticism as in the eLife assessment of eLife-RP-RA-2024-99340 (https://elifesciences.org/reviewed-preprints/99340). Hence, please refer to the responses to the companion paper.

Public Reviews:

Reviewer #1 (Public Review):

The study investigates Cancer Driving Nucleotides (CDNs) using the TCGA database, finding that these recurring point mutations could greatly enhance our understanding of cancer genomics and improve personalized treatment strategies. Despite identifying 50-150 CDNs per cancer type, the research reveals that a significant number remain undiscovered, limiting current therapeutic applications, and underscoring the need for further larger-scale research.

Strengths:

The study provides a detailed examination of cancer-driving mutations at the nucleotide level, offering a more precise understanding than traditional gene-level analyses. The authors found a significant number of CDNs remain undiscovered, with only 0-2 identified per patient out of an expected 5-8, indicating that many important mutations are still missing. The study indicated that identifying more CDNs could potentially significantly impact the development of personalized cancer therapies, improving patient outcomes.

Weaknesses:

The study is constrained by relatively small sample sizes for each cancer type, which reduces the statistical power and robustness of the findings. ICGC and other large-scale WGS datasets are publicly available but were not included in this study.

Thanks. We indeed have used all public data, including GENIE (figure 7 of the companion paper), ICGC and other integrated resources such as COSMIC. The main study is based on TCGA because it is unbiased for estimating the probability of CDN occurrences. In many datasets, the numerators are given but the denominators are not (the number of patients with the mutation / the total number of patients surveyed). In GENIE, we observed that E(u) estimated upon given sequencing panels are much smaller than in TCGA, this might be due to the selective report of nonsynonymous mutations for synonymous mutations are generally considered irrelevant in tumorigenesis.

To be able to identify rare driver mutations, more samples are needed to improve the statistical power, which is well-known in cancer research. The challenges in direct functional testing of CDNs due to the complexity of tumor evolution and unknown mutation combinations limit the practical applicability of the findings.

We fully agree. We now add a few sentences, making clear that the theory allows us to see how much more can be gained by each stepwise increase in sample size. For example, when the sample size reaches 106, further increases will yield almost no gain in confidence of CDNs identified (see figures of eLife-RP-RA-2024-99340. As pointed out in our provisional responses, an important strength of this pair of studies is that the results are testable. The complexity is the combination of mutations required for tumorigenesis and the identification of such combinations is the main goal and strength of this pair of studies. We add a few sentences to this effect.

While the importance of large sample sizes in identifying cancer drivers is well-recognized, the analytical framework presented in the companion paper (https://elifesciences.org/reviewed-preprints/99340) goes a step further by quantitatively elucidating the relationship between sample size and the resolution of CDN detection.

The question is very general as it is about multigene interactions, or epistasis. The challenges are true in all aspects of evolutionary biology, for example, the genetics of reproductive isolation(Wu and Ting 2004). The issue of epistasis is difficult because most, if not all, of the underlying mutations have to be identified in order to carry out functional tests. While the full identification is rarely feasible, it is precisely the objective of the CDN project. When the sample size increases to 100,000 for a cancer type, all point mutations for that cancer type should be identifiable.

The QC of the TCGA data was not very strict, i.e, "patients with more than 3000 coding region point mutations were filtered out as potential hypermutator phenotypes", it would be better to remove patients beyond +/- 3*S.D from the mean number of mutations for each cancer type. Given some point mutations with >3 hits in the TCGA dataset, they were just false positive mutation callings, particularly in the large repeat regions in the human genome.

Thanks. The GDC data portal offers data calls from multiple pipelines, enabling us to select mutations detected by at least two pipelines. While including patients with hypermutator phenotypes could introduce potential noise, as shown in Eq. 10 of the main text, our method for defining the upper limit of i* is relative robust to the fluctuations in the E(u) of the corresponding cancer population. Since readers may often ask about this, we expand the Methods section somewhat to emphasize this point.

The codes for the statistical calculation (i.e., calculation of Ai_e, et al) are not publicly available, which makes the findings hard to be replicated.

We have now updated the section of “Data Availability” in both papers. The key scripts for generating the major results are available at: https://gitlab.com/ultramicroevo/cdn_v1.

Reviewer #2 (Public Review):

Summary:

The study proposes that many cancer driver mutations are not yet identified but could be identified if they harbor recurrent SNVs. The paper leverages the analysis from Paper #1 that used quantitative analysis to demonstrate that SNVs or CDNs seen 3 or more times are more likely to occur due to selection (ie a driver mutation) than they are to occur by chance or random mutation.

Strengths:

Empirically, mutation frequency is an excellent marker of a driver gene because canonical driver mutations typically have recurrent SNVs. Using the TCGA database, the paper illustrates that CDNs can identify canonical driver mutations (Figure 3) and that most CDNs are likely to disrupt protein function (Figure 2). In addition, CDNs can be shared between cancer types (Figure 4).

Weaknesses:

Driver alteration validation is difficult, with disagreements on what defines a driver mutation, and how many driver mutations are present in a cancer. The value proposed by the authors is that the identification of all driver genes can facilitate the design of patient-specific targeting therapies, but most targeted therapies are already directed towards known driver genes. There is an incomplete discussion of oncogenes (where activating mutations tend to target a single amino acid or repeat) and tumor suppressor genes (where inactivating mutations may be more spread across the gene). Other alterations (epigenetic, indels, translocations, CNVs) would be missed by this type of analysis.

The above paragraph has three distinct points. We shall respond one by one.

First, … can facilitate the design of patient-specific targeting therapies, but most targeted therapies are already directed towards known driver genes…

We state in the text of Discussion the following that shows only a few best-known driving mutations have been targeted. It is accurate to say that < 5% of CDNs we have identified are on the current targeting list. Furthermore, this list we have compiled is < 10% of what we expect to find.

Direct functional test of CDNs would be to introduce putative cancer-driving mutations and observe the evolution of tumors. Such a task of introducing multiple mutations that are collectively needed to drive tumorigenesis has been done only recently, and only for the best-known cancer driving mutations (Ortmann et al. 2015; Takeda et al. 2015; Hodis et al. 2022). In most tumors, the correct combination of mutations needed is not known. Clearly, CDNs, with their strong tumorigenic strength, are suitable candidates.

Second, “There is an incomplete discussion of oncogenes (where activating mutations tend to target a single amino acid or repeat) and tumor suppressor genes (where inactivating mutations may be more spread across the gene).”

We sincerely thank the reviewer for this insightful comment. Below are two new paragraphs in the Discussion pertaining to the point:

In this context, we should comment on the feasibility of targeting CDNs that may occur in either oncogenes (ONCs) or tumor suppressor genes (TSGs). It is generally accepted that ONCs drive tumorigenesis thanks to the gain-of-function (GOF) mutations whereas TSGs derive their tumorigenic powers by loss-of-function (LOF) mutations. It is worthwhile to point out that, since LOF mutations are likely to be more widespread on a gene, CDNs are biased toward GOF mutations. The often even distribution of non-sense mutations along the length of TSGs provide such evidence. As gene targeting aims to diminish gene functions, GOF mutations are perceived to be targetable whereas LOF mutations are not. By extension, ONCs should be targetable but TSGs are not. This last assertion is not true because mutations on TSGs may often be of the GOF kind as well.

The data often suggest that mis-sense mutations on TSGs are of the GOF kind. If mis-sense mutations are far more prevalent than nonsense mutations in tumors, the mis-sense mutations cannot possibly be LOF mutations. (After all, it is not possible to lose more functions than nonsense mutations.) For example, AAA to AAC (K to Q) is a mis-sense mutation while AAA to AAT (K to stop) is a non-sense mutation. In a separate study (referred to as the escape-route analysis), we found many cases where the mis-sense mutations on TSGs are more prevalent (> 10X) than nonsense mutations. Another well-known example is the distribution of non-sense mutations TSGs. For example, on APC, a prominent TSG, non-sense mutations are far more common in the middle 20% of the gene than the rest (Zhang and Shay 2017; Erazo-Oliveras et al. 2023). The pattern suggests that even these non-sense mutations could have GOF properties.

The following response is about the clinical implications of our CDN analysis. Canonical targeted therapy often relies on the Tyrosine Kinase Inhibitors (TKIs) (Dang et al. 2017; Danesi et al. 2021; Waarts et al. 2022). Theoretically, any intervention that suppresses the expression of gain-of-function (GOF) CDNs could potentially have therapeutic value in cancer treatment. This leads us to a discussion of oncogenes versus TSGs in the context of GOF / LOF (loss of function) mutations. Not all mutations on oncogenes have oncogenic effect, besides, truncated mutations in oncogenes are often subject to negative selection (Bányai et al. 2021), the identification of CDNs within oncogenes is therefore crucial for developing effective cancer treatment guidelines. Secondly, while TSGs are generally believed to promote cancer development via loss of function mutations, research suggests that certain mutations within TSGs can have GOF-like effect, such as the dominant negative effect of truncated TP53 mutations (Marutani et al. 1999; de Vries et al. 2002; Gerasimavicius et al. 2022). Characterizing driver mutations as GOF or LOF mutations could potentially expand the scope of targeted cancer therapy. We’ll address this issue in a third study in preparation.

The method could be more valuable when applied to the noncoding genome, where driver mutations in promoters or enhancers are relatively rare, or as yet to be discovered. Increasingly more cancers have had whole genome sequencing. Compared to WES, criteria for driver mutations in noncoding regions are less clear, and this method could potentially provide new noncoding driver CDNs. Observing the same mutation in more than one cancer specimen is empirically unusual, and the authors provide a solid quantitative analysis that indicates many recurrent mutations are likely to be cancer-driver mutations.

Again, we are grateful for the comments which prompt us to expand a paragraph in Discussion, reproduced below.

The CDN approach has two additional applications. First, it can be used to find CDNs in non-coding regions. Although the number of whole genome sequences at present is still insufficient for systematic CDN detection, the preliminary analysis suggests that the density of CDNs in non-coding regions is orders of magnitude lower than in coding regions. Second, CDNs can also be used in cancer screening with the advantage of efficiency as the targeted mutations are fewer. For the same reason, the false negative rate should be much lower too. Indeed, the false positive rate should be far lower than the gene-based screen which often shows a false positive rate of >50% (supplement File S1).

Again, we are grateful that Reviewer #2 have addressed the potential value of our study in finding cancer drivers in non-coding regions. A major challenge in this area lies in defining the appropriate L value as presented in Eq. 10. In the main text, we used a gamma distribution to account for the variability of mutation rates across sites in coding region. For the non-coding region, we will categorize these regions based on biological annotations. The goal is to set different i* cutoffs for different genomic regions (such as heterochromatin / euchromatin, GC-rich regions or centromeric regions), and avoid false positive calls for CDN in repeated regions (Elliott and Larsson 2021; Peña et al. 2023).

References

Bányai L, Trexler M, Kerekes K, Csuka O, Patthy L. 2021. Use of signals of positive and negative selection to distinguish cancer genes and passenger genes. Elife 10:e59629.

Danesi R, Fogli S, Indraccolo S, Del Re M, Dei Tos AP, Leoncini L, Antonuzzo L, Bonanno L, Guarneri V, Pierini A, et al. 2021. Druggable targets meet oncogenic drivers: opportunities and limitations of target-based classification of tumors and the role of Molecular Tumor Boards. ESMO Open 6:100040.

Dang CV, Reddy EP, Shokat KM, Soucek L. 2017. Drugging the “undruggable” cancer targets. Nat Rev Cancer 17:502–508.

Elliott K, Larsson E. 2021. Non-coding driver mutations in human cancer. Nat Rev Cancer 21:500–509.

Erazo-Oliveras A, Muñoz-Vega M, Mlih M, Thiriveedi V, Salinas ML, Rivera-Rodríguez JM, Kim E, Wright RC, Wang X, Landrock KK, et al. 2023. Mutant APC reshapes Wnt signaling plasma membrane nanodomains by altering cholesterol levels via oncogenic β-catenin. Nat Commun 14:4342.

Gerasimavicius L, Livesey BJ, Marsh JA. 2022. Loss-of-function, gain-of-function and dominant-negative mutations have profoundly different effects on protein structure. Nat Commun 13:3895.

Hodis E, Triglia ET, Kwon JYH, Biancalani T, Zakka LR, Parkar S, Hütter J-C, Buffoni L, Delorey TM, Phillips D, et al. 2022. Stepwise-edited, human melanoma models reveal mutations’ effect on tumor and microenvironment. Science 376:eabi8175.

Marutani M, Tonoki H, Tada M, Takahashi M, Kashiwazaki H, Hida Y, Hamada J, Asaka M, Moriuchi T. 1999. Dominant-negative mutations of the tumor suppressor p53 relating to early onset of glioblastoma multiforme. Cancer Res 59:4765–4769.

Ortmann CA, Kent DG, Nangalia J, Silber Y, Wedge DC, Grinfeld J, Baxter EJ, Massie CE, Papaemmanuil E, Menon S, et al. 2015. Effect of Mutation Order on Myeloproliferative Neoplasms. N Engl J Med 372:601–612.

Peña MV de la, Summanen PAM, Liukkonen M, Kronholm I. 2023. Chromatin structure influences rate and spectrum of spontaneous mutations in Neurospora crassa. Genome Res. 33:599–611.

Takeda H, Wei Z, Koso H, Rust AG, Yew CCK, Mann MB, Ward JM, Adams DJ, Copeland NG, Jenkins NA. 2015. Transposon mutagenesis identifies genes and evolutionary forces driving gastrointestinal tract tumor progression. Nat Genet 47:142–150.

de Vries A, Flores ER, Miranda B, Hsieh H-M, van Oostrom CThM, Sage J, Jacks T. 2002. Targeted point mutations of p53 lead to dominant-negative inhibition of wild-type p53 function. Proceedings of the National Academy of Sciences 99:2948–2953.

Waarts MR, Stonestrom AJ, Park YC, Levine RL. 2022. Targeting mutations in cancer. J Clin Invest 132:e154943.

Wu C-I, Ting C-T. 2004. Genes and speciation. Nat Rev Genet 5:114–122.

Zhang L, Shay JW. 2017. Multiple Roles of APC and its Therapeutic Implications in Colorectal Cancer. JNCI: Journal of the National Cancer Institute 109:djw332.

Version 2

Abstract

Abstract

A central goal of cancer genomics is to identify, in each patient, all the cancer driving mutations. Among them, point mutations are referred to as Cancer Driving Nucleotides (CDNs), which recur in cancers. The companion study shows that the probability of i recurrent hits in n patients would decrease exponentially with i; hence, any mutation with i ≥ 3 hits in the TCGA database is a high-probability CDN. This study characterizes the 50∼150 CDNs identifiable for each cancer type of TCGA (while anticipating 10 times more undiscovered ones) as follows: i) CDNs tend to code for amino acids of divergent chemical properties. ii) At the genic level, far more CDNs (>5-fold) fall on non-canonical than canonical cancer driving genes (CDGs). Most undiscovered CDNs are expected to be on unknown CDGs. iii) CDNs tend to be more widely shared among cancer types than canonical CDGs, mainly because of the higher resolution at the nucleotide than the whole-gene level. iv) Most important, among the 50∼100 coding region mutations carried by a cancer patient, 5∼8 CDNs are expected but only 0∼2 CDNs have been identified at present. This low level of identification has hampered functional test and gene targeted therapy. We show that, by expanding the sample size to 105, most CDNs can be identified. Full CDN identification will then facilitate the design of patient-specific targeting against multiple CDN-harboring genes.

Introduction

Tumorigenesis in each patient is driven by mutations in the patient’s genome. Hence, a central goal of cancer genomics is to identify all driving mutations in each patient. This task is particularly challenging because each driving mutation is present in only a small fraction of patients. As the number of driver mutations in each patient has been estimated to be >5 (Armitage and Doll 1954; Bozic et al. 2010; Hanahan and Weinberg 2011; Belikov 2017; Anandakrishnan et al. 2019), the total number of driver mutations summed over all patients must be quite high.

This study, together with the companion paper (Zhang et al. 2024), are based on one simple premise: In the massively repeated evolution of cancers, any advantageous cancer-driving mutation should recur frequently, say i times in n patients. The converse that non-recurrent mutations are not advantageous is part of the same premise. We focus on point mutations, referred to as Cancer Driving Nucleotides (CDNs), and formulate the maximum of i (denoted i*) in n patients if mutations are not advantageous. For example, in the TCGA database with n generally in the range of 500∼1000, i* = 3. Hence, any point mutation with i ≥ 3 is a CDN. At present, a CDN would have a prevalence of 0.3% among cancer patients. If the sample size approaches 106, a CDN only needs to be prevalent at 5×10-5, the theoretical limit (Zhang et al. 2024).

Although there are many other driver mutations (e.g., fusion genes, chromosomal aberrations, epigenetic changes, etc.), CDNs should be sufficiently numerous and quantifiable to lead to innovations in functional tests and treatment strategies. Given the current sample sizes of various databases (Cerami et al. 2012; Weinstein et al. 2013; Tate et al. 2019; de Bruijn et al. 2023), each cancer type has yielded 50∼150 CDNs while the CDNs to be discovered should be at least 10 times more numerous. The number of CDNs currently observed in each patient is 0∼2 for most cancer types. This low-level of discovery has limited functional studies and hampered treatment strategies.

While we are proposing the scale-up of sample size to discover most CDNs, we now characterize CDNs that have been discovered. The main issues are the distributions of CDNs among genes, across cancer types and, most important, among patients. In this context, cancer driver genes (CDGs) would be a generic term. We shall use “canonical CDGs” (or conventional CDGs) for the driver genes in the union set of three commonly used lists (Bailey et al. 2018; Sondka et al. 2018; Martínez-Jiménez et al. 2020). In parallel, CDN-harboring genes, referred to “CDN genes”, constitute a new and expanded class of CDGs.

The first issue is that CDNs are not evenly distributed among genes. The canonical cancer drivers such as TP53, KRAS and EGFR tend to have many CDNs. However, the majority of CDNs, especially those yet-to-be-identified ones, may be rather evenly distributed with each gene harboring only 1∼2 CDNs. Hence, the number of genes with tumorigenic potential may be far larger than realized so far. The second issue is the distribution of CDNs and CDGs among cancer types. It is generally understood that the canonical CDGs are not widely shared among cancer types. However, much (but not all) of the presumed cancer-type specificity may be due to low statistical resolution at the genic level.

The third issue concerns the distribution of CDNs among patients. Clearly, the CDN load of a patient is crucial in diagnosis and treatment. However, the conventional diagnosis at the gene level may have two potential problems. One is that many CDNs do not fall in canonical CDGs as signals from one or two CDNs get diluted. Second, a canonical CDG, when mutated, may be mutated at a non-CDN site. In those patients, the said CDG does not drive tumorigenesis. We shall clarify the relationships between CDN mutations and genes that may or may not harbor them.

The characterizations of discovered CDNs are informative and offer a road map for expanding the CDN list. A complete CDN list for each cancer type will be most useful in functional test, diagnosis and treatment. A full list of mutations that drive the evolution of complex traits is at the center of evolutionary genetics. Such phenomena as complex human diseases (e.g., diabetes.) (Vujkovic et al. 2020; Lagou et al. 2023; Xue et al. 2023; Suzuki et al. 2024), the genetics of speciation (Q. Chen et al. 2022; Wang et al. 2022; Wu 2022) and the evolution of viruses in epidemics (Deng et al. 2022; Ruan et al. 2022; Cao et al. 2023; Ruan et al. 2023) are all prime examples in need of a full list). Thanks to their massively repeated evolution, cancers could be the first complex systems well resolved at the genic level.

Results

In molecular evolution, a gene under positive selection is recognized by its elevated evolutionary rate (Fig. 1A and 1C). There have been numerous methods for determining the extent of rate elevation (Li et al. 1985; Nei and Gojobori 1986; Yang and Swanson 2002; Lawrence et al. 2013; Martincorena et al. 2017; Pan et al. 2022; Sherman et al. 2022; Wang et al. 2022; Ruan et al. 2023) and cancer evolution studies have adopted many of them. However, no model has been developed to take advantage of the massively repeated evolution of cancers (Fig. 1B), which happens in tens of millions of people at any time.

Mutations in organismal evolution vs. cancer evolution. (A, B) A hypothetical example of DNA sequence evolution in organism vs. in cancer with the same number of mutations. (C) Mutation distribution in two species in the organismal evolution of A. (D and E) Mutation distribution in cancer evolution among 10 sequences may have D and E patterns. (F) Another pattern of mutation distribution in cancer evolution with a recurrent site but shows too few total mutations. Mutations of (F) are CDNs missed in the conventional screens.

In the whole-gene analysis, Fig. 1C-E are identical, each with A: S = 10: 1 where A and S denote nonsynonymous and synonymous mutations, respectively. However, the presence of a 4-hit site in Fig. 1E is far less likely to be neutral than Fig. 1C and 1D. Although ratio in Fig. 1F, A: S = 4: 1, is statistically indistinguishable from the neutral ratio of about 2.5: 1, Fig. 1F in fact has much more power to reject the neutral ratio than Fig. 1C and 1D. After all, the probability that multiple hits are at the same site in a big genome is obviously very small.

The analyses of CDNs across the whole genome

For the entire coding regions in the cancer genome data, we define Ai (or Si) as the number of nonsynonymous (or synonymous) sites that harbor a mutation with i recurrences. Table 1 presents the distribution of Ai and Si across the 12 cancer types with n > 300 (Weinstein et al. 2013).

For neutral mutations, we define i* as the threshold above which the expected numbers of Ai would be <1, i.e. E[Ai≥i*] 1, The corollary is that all Ai≥i* sites are advantageous CDNs. (Since Si is ∼ Ai /2.3, the same i* would apply to Si as well: E[Si≥i*] < 1,As i* is a function of the number of patients (n), it is shown mathematically in the companion study (Zhang et al. 2024) that i* = 3 for n < 1000. Interestingly, while the E[Ai≥3] is < 1, the expected E[Ai≥4] is ZZZZ, in the order of 0.001. Hence, i* = 4 may be considered unnecessarily stringent.

We should note that this study is constrained by n < 1000 in the TCGA databases. (Databases with larger n’s are also used where the actual n’s are often uncertain.) At i* = 3, we could detect only a fraction (<10%; see below) of CDNs. Many more tumorigenic mutations may be found in the i = 1 or 2 classes although not every one of them is a CDN. Since these two classes of mutations are far more numerous, they should account for the bulk of CDNs to be discovered. Indeed, Table1 shows 76 Ai≥3 CDN mutations per cancer type but 681 A2 and 56,648 A1 mutations in the lower recurrence groups. If n reaches 105∼6, most of the undiscovered CDNs in the A1 and A2 classes should be identified (Zhang et al. 2024).

In Table 2, we estimate the proportion of the A1 and A2 mutations that are possible CDNs. The relationships of A3/S3 > A2/S2, A2/S2 > A1/S1, and A1/S1 > A0/S0 are almost always observed in Table 1 with 32 (3 × 8 + 2 × 4) out of 36 such relationships. The use of A/S ratios may still under-estimate the selective advantages of A1∼3 mutations because S1∼3 may have slight advantages as well (Zhang et al. 2024). Assuming S1 is truly neutral, we use S0 to S1 as the basis to calculate the excess of A1∼3 in Table 2 where 35 of the 36 Obs(Ai) > Exp (Ai) relationships can be observed. The implication is that hundreds and, likely low thousands, of A1’s and A2’ should be CDNs whereas we have only confidently identified ∼76 strong CDNs, on average, for a cancer type. (Note that A1 excesses are less reliable since a 1% error in the calculation would mean 566 CDNs.)

CDNs and the amino acids affected

We now ask whether the amino acid changes associated with CDNs bear the signatures of positive selection. Amino acids that have divergent physico-chemical properties have been shown to be under strong selection, both positive and negative (Chen, He, et al. 2019; Chen, Lan, et al. 2019; Q. Chen et al. 2022). We note that, in almost all cases in cancer evolution, when a codon is altered, only one nucleotide of the triplet codon is changed. Among the 190 amino acid (AA, 20×19/2) pairs, only 75 of the pairs differ by one bp (Tang et al. 2004). For example, Pro (CCN) and Ala (GCN) may differ by only one bp but Pro and Gly (GGN) must differ by at least 2 bp. These 75 AA changes, referred to as the elementary AA changes (Grantham 1974; Li et al. 1985; Yang et al. 2003; Meyer et al. 2021), account for almost all AA substitutions in somatic evolution.

In a series of studies (Tang et al. 2004; Chen, He, et al. 2019; Chen, Lan, et al. 2019), we have defined the physico-chemical distances between AAs of the 75 elementary pairs as ΔUi, where i = 1 to 75. ΔUi reflects 47 measures of AA differences including hydrophobicity, size, charge etc. and ranges between 0 and 1. The most similar pair, Ser and Thr, has ΔUi= 0 and the most dissimilar pair is Asp and Try with ΔUi= 1. These studies show that ΔUi, is a strong determinant of the evolutionary rates of DNA sequences and that large-step changes (i.e., large ΔUi,’s) are more acutely “recognized” by natural selection. These large-step changes are either highly deleterious or highly advantageous. Most strikingly, advantageous mutations are enriched with AA pairs of ΔUi, > 0.8 (Chen, He, et al. 2019).

To analyze the properties of CDNs, we choose 6 cancer types from Table 1 that have the largest sample sizes (n > 500) but leap over kidney since kidney cancers have unusually low CDN counts. In Fig. 2, we divide the CDNs into groups according to the number of recurrences, i. CDNs of similar i’s are merged into the same group in the descending order of i, until there are at least 10 CDNs in the group. The 6 cancer types show two clear trends: 1) the proportion of CDNs with ΔUi, > 0.8 (red color segments) increases in groups with higher recurrences; 2) in contrast, the proportion of CDNs with ΔUi < 0.4 (green segments) decreases as recurrences increase. These two trends would mean that highly recurrent CDNs tend to involve larger AA distances (ΔUi > 0.8) and similar AAs tend not to manifest strong fitness increases. In general, CDNs alter amino acids in ways that expose the changes to strong selection.

Analysis across 6 cancer types., ranging between 0 and 1 (Tang et al. 2004; Chen, He, et al. 2019), is a measure of p differences among the 20 amino acids (see the text). The most similar amino acids have near 0 and the most dissimilar ones have Each panel corresponds to one cancer type, with horizontal bar represents distribution of each recurrence group. The numbers panel are i values and on the right are the number of sites. Note that the proportion of dark red segments increases as i increases. This that mutations at high recurrence sites (larger i’s) code for amino acids that are chemically very different from the wild type.

CDNs in relation to the genes harboring them

We shall use the term “CDN genes” for genes having at least one CDN site. Since CDN genes contribute to tumorigenesis when harboring a CDN mutation, they should be considered cancer drivers as well. CDN genes have two desirable qualities for recognition as driver genes. First, CDNs are straightforward and unambiguous to define (e.g., i ≥ 3 for n < 1000). In the literature, there have been multiple definitions of cancer driver genes (Reimand and Bader 2013; Porta-Pardo and Godzik 2014; Mularoni et al. 2016; Arnedo-Pac et al. 2019), resulting in only modest overlaps among cancer gene lists (see Fig. S3). Second, the evolutionary fitness of CDN, and hence the tumorigenic potentials of CDN genes, can be computed (supplement File S1).

We now present the analyses of CDN genes, using the same 6 cancers of Fig. 2. Two types of CDN genes are shown in Table 3. Type I genes fulfill the conventional criterion of fast evolution with the whole-gene Ka/Ks (or dN/dS) significantly larger than 1 (Martincorena et al. 2017). Averaged across cancer types, Type I overlaps by 95.7% with the canonical CDG list, which is the union of three popular lists (Bailey et al. 2018; Sondka et al. 2018; Martínez-Jiménez et al. 2020). Type I genes are mostly well-known canonical CDGs (e.g., TP53, PIK3CA, and EGFR).

Type II (CDN genes) is the new class of cancer driver genes. These genes have CDNs but do not meet the conventional criteria of whole-gene analysis. Obviously, if a gene has only one or two CDNs plus some sporadic hits, the whole-gene Ka/Ks would not be significantly greater than 1. As shown in Table 3, over 80% of CDN genes have only 1∼2 CDN sites. The salient result is that Type II genes outnumber Type I genes by a ratio of 5: 1 (229: 45, column 8, Table 3). Furthermore, Type II genes overlap with the canonical CDG list by only 23%.

Type II genes represent a new class of cancer drivers that concentrate their tumorigenic strength on a small number CDN sites. They have been missed by the conventional whole-gene definition of cancer drivers. One such example is the FGFR3 gene in lung cancer. This gene of 809 codons has only 8 hits, among which one is a CDN (i = 3) in lung cancer. It is noticed solely for this CDN. In the supplemental text, we briefly annotate these new cancer driver genes for comparisons with the canonical driver genes (supplement File S1). Possible functional tests in the future can be found in Discussion.

We now briefly discuss the driver genes listed in previous studies as shown at the lower part of Table 3 (Bailey et al. 2018; Sondka et al. 2018; Martínez-Jiménez et al. 2020). From the total number of CDGs listed, it is clear that the overlaps are limited. As analyzed before (Wu et al. 2016), conventional gene lists overlap mainly by a core set of high Ka/Ks genes. This core set has not changed much as various criteria such the replication timing, expression profiles, epigenetic features are introduced. These criteria are the reasons for the many CDGs recognized by only a small subset of CDG lists. CDN genes, in contrast, can be objectively defined as CDN mutations (i recurrences in n samples) themselves are unambiguous.

Variation in CDN number and tumorigenic contribution among genes

By and large, the distribution of CDNs among genes is very uneven. Fig. 3A shows 10 genes with at least 6 CDNs whereas 87 genes have only one CDN. Two genes stand out for the number of CDNs they harbor, TP53 and PIK3CA, which also happen to be the only genes mutated in >15% of all cancer patients surveyed (Kandoth et al. 2013). Clearly, the prevalence of mutations in a gene is a function of the number of strong CDNs it harbors.

Distribution of CDNs among genes. (A) Out of 119 CDN-carrying genes (red bars), 87 have only one CDN. For the rest, TP53 possesses the most CDNs with three others having more than 10 CDNs. (B) CDN number in TP53 among patients. The dark bar represents the observed patient number with corresponding CDNs of the X-axis. The grey bar shows the expected patient distribution. Clearly, TP53 only needs to contribute one CDN to drive tumorigenesis. Hence, TP53 (and other canonical driver genes; see text), while prevalent, does not contribute disproportionately to the tumorigenesis of each patient.

Although a small number of genes have unusually high number of CDNs, these genes may not drive the tumorigenesis in proportion to their CDN numbers in individual patients. Fig. 3B shows the number of CDN mutations on TP53 that occur in any single patient. Usually, only one CDN change is observed in a patient whereas 2 or 3 CDN mutations are expected. It thus appears that CDNs on the same genes are redundant in their tumorigenic effects such that the second hit may not yield additional advantages. This pattern of disproportionally lower contribution by CDN-rich genes is true in other genes such as EGFR and KRAS. Consequently, the large number of genes with only 1 or 2 CDN sites are disproportionately important in driving the tumorigenesis of individual patients.

CDNs in relation to the cancer types - The pan-cancer properties

In the current literature, cancer driver genes (however they are defined) generally meet the statistical criteria for driver genes in only one or a few cancer types. However, genes may in fact contribute to tumorigenesis but are insufficiently prevalent to meet the statistical requirements for CDGs. Many genes are indeed marginally qualified as drivers in some tissues and barely miss the statistical cutoff in others. To see if genes that drive tumorigenesis in multiple tissues are more common than currently understood, we need to raise the sensitivity of cancer driver detection. Thus, CDNs may provide the resolution.

To test the pan-cancer driving capacity of CDNs, we define imax as the largest i values among the 12 cancer types for each CDN. The number of cancer types where the said mutation can be detected (i.e., i > 0) is designated NC12. Fig. 4 presents the relationship between the observed NC12 of each CDN against imax of that CDN. Clearly, many CDNs are observed in multiple cancer types (NC12 > 3), even though they do not qualify as a driver gene in all but a single cancer type. It happens frequently when a mutation has i > 3 in one cancer type but has i < 3 in others. One extreme example is C394 and G395 in IDH1. In CNS, both sites show i IZ 3, while in 6 other cancer types (lung, breast, large intestine, prostate, urinary tract, liver), their hits are i < 3 but > 0. Conditional on a specific site informed by a cancer type, a mutation in another cancer type should be very unlikely if the mutation is not tumorigenic in multiple tissues. Hence, the pattern in Fig. 4 is interpreted to be drivers in multiple cancer types, but with varying statistical strength.

Sharing of CDNs across cancer types. The X-axis shows imax, which is the largest i a CDN reaches among the 12 cancer types. The Y axis shows the number of cancer types whre the mutation also occurs. Each dot is a CDN and the number of dots in the cloud is given. The blue and red dots denote, respectively, mutations classified as a CDN in one or multiple cancer types. Grey dots are non-CDNs. The table in the lower panel summarizes the number of sites and the number of genes harboring these sites.

Examining Fig. 4 more carefully, we could see that CDNs with a larger imax in one cancer type are more likely to be identified as CDNs in multiple cancer types (red dots, r = 0.97, p = 9.23×10-5, Pearson’s correlation test). Of 22 sites with imax > 20, 15 are identified as CDNs (i ≥ 3) in multiple cancer types, with a median NC12 of 9. On the opposite end, 2 CDNs with imax > 20 are observed in only one cancer type (EGFR: T2573 in lung and FGFR2: C755 in endometrium cancer). The bimodal pattern suggests that a few cancer driver mutations are tissue specific whereas most others appear to have pan-cancer driving potentials.

To conclude, when a driver is observed in more than one cancer type, it is often a cancer driver in many others, but insufficiently powerful to meet the statistical criteria for driver mutations. This pan-cancer property can be seen at the higher resolution of CDN, but is often missed at the whole-gene level. Cancers of the same tissue in different patients, often reported to have divergent mutation profiles (Nik-Zainal et al. 2012; Roberts and Gordenin 2014), should be a good test of this hypothesis.

CDNs in relation to individual patients and therapeutic strategies

In previous sections, the focus is on the population of cancer patients; for example, how many in the patient population have certain mutations. We now direct the attention to individual patients. It would be necessary to pinpoint the CDN mutations in each patient in order to delineate the specific evolutionary path and to devise the treatment strategy. We shall first address the cancer driving power of CDN vs. non-CDN mutations in the same gene.

1) Efficacy of targeted therapy against CDNs vs non-CDNs

In general, a patient would have many point mutations, only a few of which are strong CDNs. We may ask whether most mutations on the canonical genes, such as EGFR, are CDNs. Presumably, synonymous, and likely many nonsynonymous, mutations on canonical genes may not be CDNs. It would be logical to hypothesize that patients whose EGFR has a CDN mutation (Group1 patients) should benefit from the gene-targeted therapy more than patients with a non-CDN mutation on the same gene (Group2 patients). In the second group, EFGR may be a non-driver of tumorigenesis.

Published data (AACR Project GENIE Consortium 2017; Choudhury et al. 2023) are re-analyzed as shown in Fig. 5. The hypothesis that patients of Group2 would not benefit as much as those of Group1 is supported by the analysis. This pattern further strengthens the underlying assumption that non-CDN mutations, even on canonical genes, are not cancer drivers.

Survival analysis of non-small cell lung cancer (NSCLC) patients based on EGFR mutation status. Patient data were retrieved from the GENIE database (https://genie.cbioportal.org/) and stratified into three groups based on EGFR mutation profiles: Group1 comprises patients with EGFR CDN mutations; Group2 includes patients with nonsynonymous mutations in EGFR that are not CDNs; The EGFRWT group consists of patients with no EGFR mutations (see methods). Patients of Group1 and Group2 received EGFR-targeted therapies in accordance with the guidelines for managing EGFR mutant NSCLC (Passaro et al. 2022; Choudhury et al. 2023). Survival analysis using the Kaplan-Meier method revealed a significantly higher survival rate for Group1 patients compared to Group2 and the EGFRWT group (p < 0.001).

2) Number of CDNs in each patient

We postulate that a full set of CDNs should be able to inform about the cause of each cancer as well as the design of gene-targeted therapy. In Table 4, the known CDNs based on TCGA are tallied. Note that only a few CDNs fall on the canonical driver genes whereas most CDNs fall on the non-conventional ones.

In most cancer types, 10%∼30% of patients, shown in the n0 row of Table 4, have no known CDNs (and >50% among breast cancer patients). Hence, the current practice is to rely on missense mutations, regardless of CDNs or non-CDNs, on the canonical genes. The CDN column vs. the gene column in Table 4 address this issue. For example, the CDN column suggests that 33% of lung cancer patients (the n0 row) would not respond well to gene-targeted therapy whereas the gene column show only 5.3%. The difference is due to a higher, and likely inflated, detection rate of candidate drivers in the gene column. We suggest that patients who have a non-CDN mutation on a driver gene would not respond to the targeted therapy against that gene, as demonstrated in Fig. 5. In the above example, 27.7% (33%∼5.3%) of patients may be subjected to the targeted treatment but may not respond well.

3) Prevalence vs. potency of CDN-bearing genes in driving tumorigenesis

The last question is the relationship between mutation prevalence and tumorigenic strength (or potency) among CDN-bearing genes. For example, when a patient is diagnosed to have 5 CDNs in 5 genes, what may be their relative contributions to the tumorigenesis? Are they equally valid candidates for targeted therapy? It would seem logical that canonical CDGs with many CDNs should be the targets. However, because these genes would contribute at most one CDN to the tumorigenesis (Fig. 3B), targeting a high prevalence gene may not yield more benefits to the patients than targeting a low prevalence gene that has a CDN.

The implication is that prevalence and potency of CDNs may not be strongly correlated. Some genes may be prevalently mutated in the patient population but, in each affected patient, these genes may not be more potent than the less prevalent genes with a CDN mutation. Potency can be tested in vitro by gene editing or in vivo by targeting treatment. In this interpretation, targeting a CDN of low prevalence (say, i = 3) may be as effective in treatment as targeting a high prevalence CDN with i = 20. The model and Table 5 present this hypothesis based on cancer hallmarks.

The hallmarks of cancer were first proposed by (Hanahan and Weinberg 2000) with several updates (Hanahan and Weinberg 2011; Hanahan 2022). Each hallmark is a cancer phenotype shown in Table 5 that lists the number of genes involved in each particular hallmark (see Methods). While each hallmark may be associated with a number of genes, many genes are also involved in multiple hallmarks. As even the highly prevalent genes would usually have at most one mutation in each patient, we assume that each gene is associated with one hallmark in each patient.

Suppose that tumorigenesis requires a mutation in most (but perhaps not all) of the hallmarks, then the number of mutation combinations would be the product of all numbers in the corresponding column. For breast cancer, it would be 8 × 12 × 4 × 11 × 2 ∼ 1.7 × 1011. In other words, the possible mutation combinations that can drive breast cancer is over a billion. Hence, two breast cancers are unlikely to have the same set of CDGs or CDNs. In this view, the prevalence of a gene would be inversely proportional to the hallmark gene number. For example, genes of “invasion and metastasis” in breast cancer would have a prevalence of < 1/52. In contrast, the potency in tumorigenesis should depend on the hallmark phenotype itself, and independent of gene number for that hallmark. In this example, each gene of “invasion and metastasis “ may be lowly prevalent, but could also be highly potent in each patient.

In short, the prevalence and potency of CDNs may be poorly correlated. The hypothesis can be functionally tested (by gene-editing in vitro or targeting treatment in vivo) in conjunction with the data on the attraction (i.e., co-occurrences) vs repulsion (lack of co-occurrences) of CDNs.

Discussion

The companion study presents the theory that computes the limit of recurrences (i/n, i times in n patients) of reachable by neutral mutations. Above the cutoff (e.g., 3/1000), a recurrent mutation is deemed an advantageous CDN (Zhang et al. 2024). At present, the power of CDN analysis is hampered by the still small sample sizes, generally between 300 and 3000. We show that, when n reaches 105, a mutation only has to recur 12 times to be shown as a CDN, i.e., 25 times more sensitive than 3/1000. In short, nearly all CDNs should be discovered with n ≥ 105.

In this study, we apply the theory on existing data to characterize the discovered CDNs. Based on the TCGA data, this study concludes that each cancer patient carries only 1∼2 CDNs, whereas 6∼10 drivers are usually hypothesized to be present in each cancer genome (Hanahan and Weinberg 2011; Vogelstein et al. 2013; Campbell et al. 2020). This deficit signifies the current incomplete understanding of cancer driving potentials. Across patients of the same cancer type, about 50 to 150 CDNs have been discovered for each cancer type, representing perhaps only 10% of all possible CDNs. Given a complete set of CDNs, it should be possible to delineate the path of tumor evolution for each individual patient.

Direct functional test of CDNs would be to introduce putative cancer-driving mutations and observe the evolution of tumors. Such a task of introducing multiple mutations that are collectively needed to drive tumorigenesis has been done only recently, and only for the best-known cancer driving mutations (Ortmann et al. 2015; Takeda et al. 2015; Hodis et al. 2022). In most tumors, the correct combination of mutations needed is not known. Clearly, CDNs, with their strong tumorigenic strength, are suitable candidates.

Many CDNs in a patient may not fall on conventional CDGs, whereas these conventional CDGs may have passenger or weak mutations. Therefore, the efforts in gene-targeting therapy may well be shifted to the CDN-harboring genes. Given a complete set of CDNs, many more driver genes can be identified. Since many driver genes cannot be targeted for biological or technical reasons (Dang et al. 2017; Danesi et al. 2021; Waarts et al. 2022), a large set of CDGs will be desirable. The goal is that each cancer patient would have multiple targetable CDGs, all driven by CDNs they carry. In that case, the probability that resistance mutations eluding multiple targeting drugs should be diminished (B. Chen et al. 2022; Zhai et al. 2022; Bian et al. 2023; Lin et al. 2023; Zhu et al. 2023).

In this context, we should comment on the feasibility of targeting CDNs that may occur in either oncogenes (ONCs) or tumor suppressor genes (TSGs). It is generally accepted that ONCs drive tumorigenesis thanks to the gain-of-function (GOF) mutations whereas TSGs derive their tumorigenic powers by loss-of-function (LOF) mutations. Nevertheless, since LOF mutations are likely to be widespread on TSGs, they are less likely to recur as CDNs. The even distributions of non-sense mutations along the length of many TSGs provide such evidence. Importantly, as gene targeting aims to diminish gene functions, GOF mutations are perceived to be targetable whereas LOF mutations are not. By extension, ONCs should be targetable but TSGs are not, an assertion we address below.

The data suggest that mis-sense mutations on TSGs may often be of the GOF kind. If mis-sense mutations are far more prevalent than nonsense mutations in tumors, the mis-sense mutations cannot possibly be LOF mutations. (After all, it is not possible to lose more functions than nonsense mutations.) In a separate study (Deng et al. in prep.), we compare mis-sense and nonsense mutations (referred to as the escape-route analysis). For example, AAA to AAC (K to Q) is a mis-sense mutation while the same AAA codon to AAT (K to stop) is a non-sense mutation. We found many cases where the mis-sense mutations on TSGs are more prevalent (> 10X) than nonsense mutations. We interpret these mis-sense mutations to be of the GOF kind because they could not possibly “lose” more functions than the nonsense mutations Another interesting pattern may be the distributions of CDNs across different cancer types. Cancer evolution in different tissues represents parallel evolution driven by similar selection for cell proliferation but under different ecological conditions. Fig. 4 suggests that CDNs previously identified to be cancer-specific may have pan-cancer effects. In different cancer types, the same CDNs may drive the tumorigenesis but the strength may not be sufficient to raise the data above the statistical threshold.

The CDN approach has two additional applications. First, it can be used to find CDNs in non-coding regions. Although the number of whole genome sequences at present is still insufficient for systematic CDN detection, the preliminary analysis suggests that the density of CDNs in non-coding regions is orders of magnitude lower than in coding regions. Second, CDNs can also be used in cancer screening with the advantage of efficiency as the targeted mutations are fewer. For the same reason, the false negative rate should be much lower too. Indeed, the false positive rate should be far lower than the gene-based screen which often shows a false positive rate of >50% (supplement File S1).

Cancer evolution falls within the realm of ultra-microevolution (Wu et al. 2016). The repeated evolution addresses the single most severe criticism of evolutionary studies, namely all evolutionary events have a sample size of one. Such repeated evolution offers the opportunity to uncover the full list of mutations underling complex traits that is at the heart of molecular evolution. The genetics of speciation (Wu and Ting 2004; Pan et al. 2022; Wang et al. 2022; Wu 2023) and the emergence of major viral strains (such as COVID-19) (Deng et al. 2022; Ruan et al. 2022; Cao et al. 2023; Ruan et al. 2023) are both phenomena of complex gene interactions. The two companion studies may thus unite evolutionary biology and cancer medicine.